---
title: '6a: Random intercepts and random slopes'
author: "Eser Comak"
date: "December 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Preparations
```{r cars}
#install.packages(c("mgcv", "itsadug", "colorspace"), repos="https://cloud.r-project.org")
#load packages
library(mgcv)
library(itsadug)
library(colorspace)
library(plyr)
#load data
rm(list=ls())

# load data:
load("C:/Users/eserc/Desktop/HMC/second year/Ib/Advanced statistical modelling/GAMM/lab6a/dat.rda")
head(dat)
```


```{r}
 #AVERAGES:
avg <- ddply(dat, c("Timebin", "Length", "PairType", "Fan"), summarise,
             F3 = mean(F3, na.rm=TRUE))


# SETUP EMPTY PLOT WINDOW:
plot(c(), c(),
     xlim=range(avg$Timebin), ylim=range(avg$F3), 
     main="Target", ylab="F3 Amplitude", xlab="Time (ms)",
     bty='n')
abline(h=0, v=0, lty=c(1,3))

# ADD LINES FOR CONDITIONS:
with(avg[avg$Length=="Short" & avg$PairType=="Target" & avg$Fan==1,],
  lines(Timebin, F3, type='l', lwd=1.5))
with(avg[avg$Length=="Short" & avg$PairType=="Target" & avg$Fan==2,],
  lines(Timebin, F3, type='l', col=2, lwd=1.5))
with(avg[avg$Length=="Long" & avg$PairType=="Target" & avg$Fan==1,],
  lines(Timebin, F3, type='l', lty=4))
with(avg[avg$Length=="Long" & avg$PairType=="Target" & avg$Fan==2,],
  lines(Timebin, F3, type='l', col=2, lty=4))

# LEGEND:
legend_margin('topright', legend=c("Fan 1, short", "Fan 2, short", "Fan 1, long", "Fan 2, long"),
              col=rep(1:2, 2), lwd=rep(c(1.5,1), each=2), lty=rep(c(1,4), each=2), bty='n')
```

```{r}
#NOTE: EEG amplitudes are conventionally plotted with the negative values upwards and positive values downwards. To reverse the y-axis, just re-arrange the ylim values as illustrated below.
plot(c(), c(),
     xlim=range(avg$Timebin), ylim=rev(range(avg$F3)), 
     main="Target", ylab="F3 Amplitude", xlab="Time (ms)",
     bty='n')
abline(h=0, v=0, lty=c(1,3))

# ADD LINES FOR CONDITIONS:
with(avg[avg$Length=="Short" & avg$PairType=="Target" & avg$Fan==1,],
  lines(Timebin, F3, type='l', lwd=1.5))
with(avg[avg$Length=="Short" & avg$PairType=="Target" & avg$Fan==2,],
  lines(Timebin, F3, type='l', col=2, lwd=1.5))
with(avg[avg$Length=="Long" & avg$PairType=="Target" & avg$Fan==1,],
  lines(Timebin, F3, type='l', lty=4))
with(avg[avg$Length=="Long" & avg$PairType=="Target" & avg$Fan==2,],
  lines(Timebin, F3, type='l', col=2, lty=4))

# LEGEND:
legend_margin('topright', legend=c("Fan 1, short", "Fan 2, short", "Fan 1, long", "Fan 2, long"),
              col=rep(1:2, 2), lwd=rep(c(1.5,1), each=2), lty=rep(c(1,4), each=2), bty='n')
```




```{r}
#order data
dat <- dat[order(dat$Subject,dat$Length,dat$PairType,dat$Fan,dat$Timebin), ]
```


##Part I. Random effects
#Q1. 1-D smooths
First, we would like to investigate whether there is a difference in ERP response to the type of pair (Target, New Foil and Re-paired Foil). Setup a GAM model that includes an interaction between PairType and Timebin. Use the channel F3 as dependent variable.
```{r}
#make pairtype categorical again
dat$PairType <-as.factor(dat$PairType)

#model
m1 <- bam(F3 ~ PairType + s(Timebin, by = PairType) , data=dat, discrete = TRUE)
summary(m1)
```

Inspect the summary.

**What does the intercept represent?  **
mean ERP value when Timebin is at 0 and pair tye is new.

**Which estimates do we need to sum up for plotting the regression line of PairType RPFoil?  **
-0.61958 + 0.09685 + 8.932
**What can we conclude from the smooth term summary about the differences between the three PairType conditions?  **
the intteraction has similair effects....


Visualization of the estimated regression lines.
```{r}
#Run the following code for visualizing the summed effects
plot_smooth(m1, view="Timebin", plot_all="PairType", v0=0)
```
Describe at which moment the conditions seem to diverge.  

<ans> It looks like they start to diverge right around 600ms time mark </ans>

```{r}
#The code below could be used to visualize the differences between the conditions. Run the code:

plot_diff(m1, view="Timebin", comp=list(PairType=c("Target", "RPFoil")), v0=0)
plot_diff(m1, view="Timebin", comp=list(PairType=c("Target", "NewFoil")), v0=0)
```
What do the difference plots represent?

<ans> </ans>
```{r}
#Inspect the residuals of the model using a QQ-norm plot and a fitted vs residuals plot:
qqnorm(resid(m1))
qqline(resid(m1))

plot(fitted(m1), resid(m1))
```
What do you conclude about the model fit when you see the model's residuals?

<ans> </ans>



##Q3. Random intercepts
#Extend model m1 with random intercepts for participants. Make sure that the participant number is converted to a factor.
```{r}
#make subjct categorical again
dat$Subject <-as.factor(dat$Subject)
m2 <- bam(F3 ~ PairType + s(Timebin, by = PairType) + s(Subject, bs ="re"), data=dat, discrete = TRUE)
summary(m2)
```

Why is the estimated edf of the random effect term so large?
Do you think the random intercept for participants contributed to the model? Why (not)?

```{r}
plot_smooth(m1, view="Timebin", plot_all="PairType", v0=0)
plot_smooth(m2, view="Timebin", plot_all="PairType", v0=0, rm.ranef=TRUE)

qqnorm(resid(m2))
qqline(resid(m2))

plot(fitted(m2), resid(m2))
```

```{r}

attach(dat)
dat$Event <- interaction(dat$Subject, dat$Length, dat$PairType, dat$Fan, drop = TRUE)

#n parameter specifies th number of red lines
plot_modelfit(m2, view = "Timebin",event=dat$Event)
```

##PART 2
#interaction non linear
```{r}

attach(dat)
dat <- dat[order(Subject,Length,PairType,Fan,Timebin), ]
```
non l
#random smooths
```{r}
#EITHER random intrcepts and random slopes or random smooths

attach(dat)
# fs is factor smooth
m4<-bam(F3~PairType + s(Timebin, by=PairType) +s(Timebin, Subject, bs="fs", m=1),data = dat, discrete = TRUE)
plot(m4)

#modl fits
plot_modelfit(m4, view = "Timebin",event=dat$Event)

summary(m4)
```